\Ace: Growing a Statically-Typed Language Inside a Python

### 2013-11-15 ace-pldi14/ace-pldi14.tex
pldi14 section 2 first draft done (whew)

The type \verb|TF| on line \ref{compscript}.5 is extracted from the generic function \verb|negate| constructed in Listing \ref{metaprogramming}. Generic functions, according to Sec. \ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless extracting a type from it. Although a conventional arrow type cannot be assigned to \verb|negate|, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \verb|negate|''. This type could also have been explicitly written as \verb|ace.GenericFnType(listing2.negate)|. During typechecking and translation of \verb|map_neg_f32| and \verb|map_neg_ci32|, the call to \verb|f| on line \ref{map}.6 operates by using the types of the provided arguments to compile the generic function that inhabits the singleton type of \verb|f| (\verb|negate| in both of these cases) to a concrete function. This is why there are two versions of \verb|negate| in the output in Listing \ref{mapout}. In other words, types \emph{propagate} into generic functions -- we didn't need to compile \verb|negate| explicitly. This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \verb|input|, the indexing operation on line \ref{map}.6 resulted in an error. A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \verb|clx.size_t| (the return type of the OpenCL primitive function \verb|get_global_id|, see below).

@@

\subsection{Implicit Compilation and Execution}\label{compenv}\label{backend}\label{implicit}
%Professional end-user programmers (e.g. scientists and engineers \cite{professional-end-users}) today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}

@@

Ace supports a refinement to this workflow, as an alternative to the \verb|acec| compiler described above. In other words, \verb|acec| can generate source code for kernels but it does not specify how and from what language thet will be called. This mode of use allows Python to be the calling language.

At the time of writing, this is only supported with the OpenCL backend, but other backends can implement this feature as well by satisfying a simple interface specification. The OpenCL host API wraps and is a superset of the \verb|pyopencl|, adding a simpler type-aware API. Both generic functions and concrete functions can then be called like regular Python functions, with additional keyword arguments specifying the global and local size (i.e. the number of threads and how they are grouped). Device buffers can carry type information, unlike in the basic OpenCL host API, and this type information can be propagated from Python to Ace functions directly, as if the call had been within a kernel,  eliminating the \verb|compile| step that we have been using thus far and thus all mention of types.

An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 12-13. 
The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 9-10. This wrapper associates type information with each buffer, based on the type of the \verb|numpy| array, and this is used to implicitly compile \verb|map| as appropriate the first time it is called for any given combination of input types. Notice also that \verb|add5| is passed in directly.

By way of comparison, the same program written using the OpenCL C API directly is an order of magnitude larger and significantly more difficult to read and understand. It does not support higher-order functions nor is there any way to write \verb|map| in a type-generic way. A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{klockner2011pycuda} is still twice as large and more difficult to comprehend than the code we have shown thus far. Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.

>>>>

The type \verb|TF| on line \ref{compscript}.5 is extracted from the generic function \verb|negate| constructed in Listing \ref{metaprogramming}. Generic functions, according to Sec. \ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless extracting a type from it. Although a conventional arrow type cannot be assigned to \verb|negate|, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \verb|negate|''. This type could also have been explicitly written as \verb|ace.GenericFnType(listing2.negate)|. During typechecking and translation of \verb|map_neg_f32| and \verb|map_neg_ci32|, the call to \verb|f| on line \ref{map}.6 operates by using the types of the provided arguments to compile the generic function that inhabits the singleton type of \verb|f| (\verb|negate| in both of these cases) to a concrete function. This is why there are two versions of \verb|negate| in the output in Listing \ref{mapout}. In other words, types \emph{propagate} into generic functions -- we didn't need to compile \verb|negate| explicitly. This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \verb|input|, the indexing operation on line \ref{map}.6 resulted in an error. A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \verb|clx.size_t| (the return type of the OpenCL primitive function \verb|get_global_id|).

@@

\subsection{Implicit Compilation and Interactive Execution}\label{compenv}\label{backend}\label{implicit}
%Professional end-user programmers (e.g. scientists and engineers \cite{professional-end-users}) today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}

@@

Ace supports a refinement to this workflow, as an alternative to the \verb|acec| compiler described above, for targets that have wrappers like this available, including \verb|clx.opencl|. Listing \ref{py} shows an example of this workflow where the user chooses a compute device (line \ref{py}.3),  constructs a \verb|numpy| array (line \ref{py}.5), transfers it to the device (line \ref{py}.6), allocates an empty equal-sized buffer for the result of the computation (line \ref{py}.7), launches the {generic} kernel \verb|map| from Listing \ref{map} with these device arrays as well as the function \verb|negate| from Listing \ref{metaprogramming} (line \ref{py}.9) choosing a number of threads equal to the number of elements in the input array (line \ref{py}.10), and transfers the result back into main memory to check that the device computed the expected result (line \ref{py}.12).

For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising. This simplicity is enabled largely due to  implicit uses of types throughout the code. First, \verb|numpy| keeps track of the type, shape and order of its arrays. The type of \verb|input|, for example, is \verb|numpy.double| by default, its shape is \verb|(1024,)| and its order is row-major by default. The \verb|pyopencl| library that our mechanism is built atop uses this information to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size. Our wrapper of \verb|pyopencl| retains this information in the Python representations of the device arrays, \verb|d_input| and \verb|d_output|. An Ace type parameterized can use this metadata to typecheck and generate code for operations like indexing in Ace functions\todo{rephrase}.

This allows us to call the generic function \verb|map| directly on these Python data structures (as well as the generic function \verb|negate|) without first requiring an explicit type assignment, like we needed when using \verb|acec| above. In other words, dynamic types and other metadata can propagate from Python data structures into an Ace generic function as static type information, in the same manner as it propagated \emph{between} generic functions in the previous section. In both cases, typechecking and translation of \verb|map| happens the first time a particular type assignment is encountered and cached for subsequent use. When called from Python, the generated OpenCL source code is also compiled for the device we selected using the OpenCL host API's compiler infrastructure, and cached.

The same program written using the OpenCL C API directly is an order of magnitude longer and significantly more difficult to comprehend. OpenCL not support higher-order functions nor is there any way to write \verb|map| in a type-generic manner. If we instead use the \verb|pyopencl| library and apply the techniques described in \cite{klockner2011pycuda}, the program is still twice as large and less readable than this code. Both the \verb|map| and \verb|negate| functions must be explicitly specialized with the appropriate types using string manipulation techniques. Higher order functions are still not available, and must be simulated by string manipulation. That approach also does not permit the use any of the language extensions that Ace enables (in this case, the only interesting use of the extension mechanism was by the Ace type for \verb|numpy| arrays, which ensures that indexing respects ordering; see Sec. \ref{examples} for more interesting possibilities).\todo{would be aesthetically pleasing for a line to be shaved off so this ends up on the previous page}

%Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
